Map{R}educe in {MPI} for Large-scale Graph Algorithms

### 2010-04-30 paper/graph.tex
Added improved R-MAT generating algorithm and finished PageRank algorithms.
git-svn-id: svn+ssh://software-srn.sandia.gov/svn/mapreduce/trunk@594 3fc257bb-8550-0410-81b3-a5c103da83f7

@@

Input from R-MAT generator: \\
\> MapReduce object $V$ containing all graph vertices:  Key=$v_i$, Value = NULL.  \\
\> MapReduce object $E$ containing all graph edges: Key=$v_i$, Value = $v_j$. \\
Initialize MapReduce object for matrix $A$  and \\
MapReduce object $MT$ containing all vertices with no outlinks.\\
\> Copy: $A = E$ \\
\> Add: $V$ to $A$ \\
\> Collate $A$: Vertex is Key \\
\> Reduce $A$:  \\
\> \> Input:  Key = $v_i$; MultiValue = NULL,$v_j$,$v_k$,... \\
\> \> If (${nvalues} > 1$) \\
\> \> \> Compute: $d_{out}(v_i) = {nvalues} - 1$. \\
\> \> \> Output:  For each $v_j$ in MultiValue, Key = $v_i$; Value = ($v_j$, $\alpha/d_{out}(v_i)$). \\
\> \> else \\
\> \> \> Output to $MT$: Key = $v_i$; Value = 0. \\
Initialize MapReduce object for pagerank vector $x$. \\
\> Map:   \\
\> \> Input:  Graph vertices $V$. \\
\> \> Output:  For each vertex, Key = $v_i$; Value = $1/|V|$. \\
While (${residual} > {tolerance}$) \\
\> Compute contribution for random jumps and vertices with no outlinks. \\
\> \> Add: $V$ to $MT$. \\
\> \> Collate $MT$:  Vertex is Key. \\
\> \> Reduce $MT$:   \\
\> \> \> Input:  Key = $v_i$; MultiValue = $x_i, 0$ if $v_i \in {MT}$; MultiValue = $x_i$ if $v_i \notin {MT}$. \\
\> \> \> Compute:  Contributions; store in data pointer. \\
\> \> \> Output:  If ${nvalues} = 2$, Key = $v_i$; Value = 0. \\
\> \> MPI\_Allreduce with MPI\_SUM to compute global contributions. \\
\> Compute $y = A^T x$.

@@

The MapReduce implementation performs two {\it map()} operations to
initialize the graph matrix $A$ and PageRank vector $x$.  A {\it
collate()} operation gathers all row entries $a_{ij}$ with their
associated $x_i$ entry, and a {\it reduce()} computes $a_{ij} x_i$.  A
second {\it collate()} gathers, for each $j$, all contributions to the
column sum $\sum a_{ij} x_i$, which is computed by a second {\it
reduce()}.  MPI\_Allreduce calls are used to compute global norms and
residuals.

>>>>

By using more than one MapReduce object, we can improve the performance
of the R-MAT generation algorithm.  In the improved algorithm
(Figure~\ref{fig:rmat2}, we emit newly created edges into MapReduce object
$E_{new}$ and aggregate the edges to processors.  We then add those edges
to the MapReduce object $E_{old}$, containing all previously generated
unique edges.  We convert the total set of key/values to key/multivalues
and use a reduce operation to remove duplicates.  Performance improvements
in this algorithm result from the reduced number of key/values in the aggregate
operation, compared to the collate operation in the original algorithm, 
particularly in later iterations of the generation where few
edges are created.

$M_{\rm remain} = M$ \\
while $M_{\rm remain} > 0$: \\
\> 1 Map $E_{new}$: \> Generate $M_{\rm remain}/P$ random edges on each processor \\
           \> \> output Key = $(V_i,V_j)$, Value = NULL \\
\> 1 Aggregate $E_{new}$. \\
\> 1 Add $E_{new}$ to $E_{old}$. \\
\> 1 Convert $E_{old}$ from KV to KMV. \\
\> 1 Reduce $E_{old}$: \> Remove duplicate edges \\
              \> \> input Key = $(V_i,V_j)$, MultiValue = one or more NULLs \\
              \> \> output Key = $(V_i,V_j)$, Value = NULL \\
\> $M_{\rm remain} = M - N_{kv}$

 \caption{Improved MapReduce algorithm for R-MAT graph generation.}

 \label{fig:rmat2}
\end{figure}

@@

%Input from R-MAT generator: \\
%\> MapReduce object $V$ containing all graph vertices:  Key=$v_i$, Value = NULL.  \\
%\> MapReduce object $E$ containing all graph edges: Key=$v_i$, Value = $v_j$. \\
%Initialize MapReduce object for matrix $A$  and \\
%MapReduce object $MT$ containing all vertices with no outlinks.\\
%\> Copy: $A = E$ \\
%\> Add: $V$ to $A$ \\
%\> Collate $A$: Vertex is Key \\
%\> Reduce $A$:  \\
%\> \> Input:  Key = $v_i$; MultiValue = NULL,$v_j$,$v_k$,... \\
%\> \> If (${nvalues} > 1$) \\
%\> \> \> Compute: $d_{out}(v_i) = {nvalues} - 1$. \\
%\> \> \> Output:  For each $v_j$ in MultiValue, Key = $v_i$; Value = ($v_j$, $\alpha/d_{out}(v_i)$). \\
%\> \> else \\
%\> \> \> Output to $MT$: Key = $v_i$; Value = 0. \\
%Initialize MapReduce object for pagerank vector $x$. \\
%\> Map:   \\
%\> \> Input:  Graph vertices $V$. \\
%\> \> Output:  For each vertex, Key = $v_i$; Value = $1/|V|$. \\
Input: \\
\> MapReduce object $A$ containing $|V| \times |V|$ matrix:  Key = $i$, Value = $[j, a_{ij}]$ \\
\> MapReduce object $MT$ containing indices of empty rows of $A$:  Key = $i$, Value = 0. \\
\> MapReduce object $x$ containing initial PageRank vector:  Key = $i$, Value = $1/|V|$. \\
While (${residual} > {tolerance}$) \\
\> Compute contribution for random jumps and zero-outdegree vertices. \\
\> \> Add: $x$ to $MT$. \\
\> \> Collate $MT$:  Vertex is Key. \\
\> \> Reduce $MT$:   \\
\> \> \> Input:  Key = $i$; MultiValue = $x_i, 0$ if $i \in {MT}$; MultiValue = $x_i$ if $i \notin {MT}$. \\
\> \> \> Compute:  Local contributions $cont$. \\
\> \> \> Output:  If ${nvalues} = 2$, Key = $v_i$; Value = 0. \\
\> \> MPI\_Allreduce with MPI\_SUM to compute global contribution $gcont$. \\
\> Compute $y = A^T x$.  \\
\> \> Copy $x$ to $y$. \\
\> \> Add $A$ to $y$. \\
\> \> Collate $y$:  Key is row index of $A$. \\
\> \> Reduce $y$: \\
\> \> \>  Input: Key = $i$; Multivalue = $i, [j, a_{ij}]$ for nonzeros $a_{ij}$ in $A$ \\
\> \> \>  Output:  Key = $j$; Value = $a_{ij} y_i$. \\
\> \> Collate $y$:  Key is column index of $A$. \\
\> \> Reduce $y$:  \\
\> \> \> Input:  Key = $j$; Multivalue = $a_{ij} y_i$ for all non-zero entries $a_{ij}$ of column $j$ \\
\> \> \> Output:  Key = $j$; Value = $\sum_i a_{ij} y_i$. \\
\> Add contribution $gcont$ to $y$; compute local max norm $norm$ of $y$. \\
\> \> Map $y$:  \\
\> \> \> Input:  Key = $i$, Value = $y_i$. \\
\> \> \> Compute: $y_i = y_i + gcont$; if $y_i > {norm}$, $norm = y_i$. \\
\> \> \> Output:  Key = $i$, Value = $y_i$. \\
\> \> MPI\_Allreduce $norm$ with MPI\_MAX to compute global max norm $gnorm$. \\
\> Scale $y$ by $gnorm$. \\
\> \> Map $y$: \\
\> \> \> Input:  Key = $i$, Value = $y_i$. \\
\> \> \> Output:  Key = $i$, Value = $y_i / {gnorm}$. \\
\> Compute residual. \\
\> \> Add $y$ to $x$. \\
\> \> Collate $x$:  Key is index $i$. \\
\> \> Reduce $x$:   \\
\> \> \> Input:  Key = $i$; Multivalue = $x_i, y_i$. \\
\> \> \> Compute:  if $|x_i - y_i| > {resid}$, ${resid} = |x_i - y_i|$ \\
\> \> MPI\_Allreduce with MPI\_MAX to compute $residual$. \\
\> $x = y$; \\
Output:  PageRank vector $x$.
  \end{tabbing}
 \end{minipage}}\end{center}

@@

%The MapReduce implementation performs two {\it map()} operations to
%initialize the graph matrix $A$ and PageRank vector $x$.  
The key kernel of PageRank is the matrix-vector multiplication $y = A^T x$.
In it, a {\it
collate()} operation gathers all row entries $a_{ij}$ with their
associated $x_i$ entry, and a {\it reduce()} computes $a_{ij} x_i$.  A
second {\it collate()} gathers, for each $j$, all contributions to the
column sum $\sum_i a_{ij} x_i$, which is computed by a second {\it
reduce()}.  MPI\_Allreduce calls are used to compute global norms and
residuals.

The entire PageRank algorithm can be made more efficient by calling
{\it aggregate()} on the MapReduce objects representing the matrix $A$, 
the empty matrix rows $MT$, and the vector $x$ once before the PageRank 
iterations begin.  Since these three MapReduce
objects have the same key types, pre-aggregating the objects moves all
Key-Value pairs with a given key to a single processor.  Thus, many of
the PageRank operations become local operations, replacing 
the {\it collate()} operation (which is equivalent to {\it aggregate()} 
followed by {\it convert()}) with only a local {\it convert()} call. 
Figure~\ref{fig:pr2} shows the
improved algorithm; modified operations are marked with an asterisk.
This strategy of pre-aggregating to improve data locality is useful in
many MapReduce algorithms, but assumes that keys are always mapped to
the same processor.

Input: \\
$*$\> Pre-aggregated MapReduce object $A$ containing $|V| \times |V|$ matrix:  Key = $i$, Value = $[j, a_{ij}]$ \\
$*$\> Pre-aggregated MapReduce object $MT$ containing indices of empty rows of $A$:  Key = $i$, Value = 0. \\
$*$\> Pre-aggregated MapReduce object $x$ containing initial PageRank vector:  Key = $i$, Value = $1/|V|$. \\
While (${residual} > {tolerance}$) \\
\> Compute contribution for random jumps and zero-outdegree vertices. \\
\> \> Add: $x$ to $MT$. \\
$*$\> \> Convert $MT$:  Vertex is Key. \\
\> \> Reduce $MT$:   \\
\> \> \> Input:  Key = $i$; MultiValue = $x_i, 0$ if $i \in {MT}$; MultiValue = $x_i$ if $i \notin {MT}$. \\
\> \> \> Compute:  Local contributions $cont$. \\
\> \> \> Output:  If ${nvalues} = 2$, Key = $v_i$; Value = 0. \\
\> \> MPI\_Allreduce with MPI\_SUM to compute global contribution $gcont$. \\
\> Compute $y = A^T x$.  \\
\> \> Copy $x$ to $y$. \\
\> \> Add $A$ to $y$. \\
$*$\> \> Convert $y$:  Key is row index of $A$. \\
\> \> Reduce $y$: \\
\> \> \>  Input: Key = $i$; Multivalue = $i, [j, a_{ij}]$ for nonzeros $a_{ij}$ in $A$ \\
\> \> \>  Output:  Key = $j$; Value = $a_{ij} y_i$. \\
\> \> Collate $y$:  Key is column index of $A$. \\
\> \> Reduce $y$:  \\
\> \> \> Input:  Key = $j$; Multivalue = $a_{ij} y_i$ for all non-zero entries $a_{ij}$ of column $j$ \\
\> \> \> Output:  Key = $j$; Value = $\sum_i a_{ij} y_i$. \\
\> Add contribution $gcont$ to $y$; compute local max norm $norm$ of $y$. \\
\> \> Map $y$:  \\
\> \> \> Input:  Key = $i$, Value = $y_i$. \\
\> \> \> Compute: $y_i = y_i + gcont$; if $y_i > {norm}$, $norm = y_i$. \\
\> \> \> Output:  Key = $i$, Value = $y_i$. \\
\> \> MPI\_Allreduce $norm$ with MPI\_MAX to compute global max norm $gnorm$. \\
\> Scale $y$ by $gnorm$. \\
\> \> Map $y$: \\
\> \> \> Input:  Key = $i$, Value = $y_i$. \\
\> \> \> Output:  Key = $i$, Value = $y_i / {gnorm}$. \\
\> Compute residual. \\
\> \> Add $y$ to $x$. \\
$*$\> \> Convert $x$:  Key is index $i$. \\
\> \> Reduce $x$:   \\
\> \> \> Input:  Key = $i$; Multivalue = $x_i, y_i$. \\
\> \> \> Compute:  if $|x_i - y_i| > {resid}$, ${resid} = |x_i - y_i|$ \\
\> \> MPI\_Allreduce with MPI\_MAX to compute $residual$. \\
\> $x = y$; \\
Output:  PageRank vector $x$.
  \end{tabbing}
 \end{minipage}}\end{center}

 \caption{Improved MapReduce algorithm for PageRank vertex ranking; pre-aggregating the MapReduce objects allows many communication intensive {\it collate()} 
operations to be replaced by with strictly local {\it convert()} operations (as marked by the asterisks).}

 \label{fig:pr2}
\end{figure}
